Application of the least-squares method to the multiscale transport equation in porous media

### 2011-05-09 MSTE_methodology.tex
some comments after review

Traditionally, the transport of a given specie with concentration $\phi$ convected by plug flow through a porous media without diffusion in one dimension (z) can be written as:

\begin{equation}
\epsilon v \frac{\phi(z)}{\partial z}=S(z)
\end{equation}

@@

\begin{align}
q(z, \xi) \cdot \frac{\partial \phi(z, \xi)}{\partial z}  = 
 S(z, \xi) + \int [\phi(z, \hat \xi) - \phi(z, \xi)] h(z, \hat \xi \rightarrow \xi) d\hat \xi . 
\label{eq:ss2}
\end{align}

The left hand side and first term of the right hand side of the equation \ref{eq:ss2} are the convection per scale and the source per scale and present no major interaction between scales. However, the integral on the right hand side couples different scales and makes impossible to solve the problem for each scale independently. It should be pointed out that advection can be happening at different speeds ($\tfrac{\partial q}{\partial \xi} \neq 0$) and the source term can also be written independently for each pore-size.

@@

Find $\mathbf{f}\in X(\Omega)$ such that:
\begin{eqnarray}
\mathcal{L} \mathbf{f} = \mathbf{g} \quad \mbox{in} \ \Omega \label{eq:Problem} \\
\mathcal{B} \mathbf{f} = \mathbf{f_\Gamma} \quad \mbox{on} \ \Gamma \label{eq:Boundary}
\end{eqnarray}

@@

\begin{eqnarray}
\mathbf{f}&=&\phi_i \nonumber \\
\mathbf{g}&=&S_i \label{eq:problem} \\
\mathcal{L}\bullet&=& q \cdot \frac{\partial \bullet}{\partial z} - \int \bullet h(z, \hat \xi \rightarrow \xi) d\hat \xi +\bullet\int h(z, \hat \xi \rightarrow \xi) d\hat \xi \nonumber
\end{eqnarray}

@@

\begin{align}
\epsilon(\xi)=&A (\xi-\xi_0) (-\xi+\xi_f) \\
q({\xi}) =&V \epsilon(\xi) \xi^2\\
h(\hat \xi, \xi, z) =& q(\hat \xi) q(\xi)/L\\
\phi(\xi, z) =& 10 sin(z) (1-M Log({\xi}/R))\\
\nonumber S(\xi, z)=& 10 A V cos(z) {\xi}^2 ({\xi}-\xi_0) (\xi_f-{\xi}) \left(1-M \text{Log}\left[\nicefrac{{\xi}}{R}\right]\right) +\\
\nonumber  &A^2 M V^2  (360 L)^{-1} {\xi}^2 ({\xi}-\xi_0) ({\xi}-\xi_f) sin(z) \times ... \\
\nonumber  &\big(-81 \xi_0^5+175 \xi_0^4 \xi_f-175 \xi_0 \xi_f^4+  ...\\
\nonumber &81 \xi_f^5+60 \xi_0^4 (3 \xi_0-5 \xi_f) \text{Log}\left[\frac{\xi_0}{{\xi}}\right]+ ...\\
& 60 (5 \xi_0-3 \xi_f) \xi_f^4 \text{Log}\left[\frac{\xi_f}{{\xi}}\right]\big)
\end{align}

@@

The solution is shown in Figure \ref{fig:solution}. As the liquid flows down the column $\phi$ increases due to reactions happening inside the pores. Despite the source term is very different for different pore sizes (negative for large pores and positive for small pores, see Figure \ref{fig:source}) the tracer $\phi$ is exchanged between different pores scales tending to level out the differences. However the increase is larger for the small pores at all the length of the column (z values). The source term (Figure \ref{fig:source}) behaves differently for different pore scales. Large amounts of $\phi$ are generated for pores smaller than 5 mm, while a sink is observed in the largest pores. This behaviour for the source term mimic two competing reactions. It models the case of a specie that is a product of an heterogeneous reaction but is a reactant of a reaction taking place in the liquid bulk. Heterogeneous reactions depend strongly of the surface to volume ratio which is much larger for small pore scales. At the same time the reactant is consumed in the larger pores in this case as the bulk reaction that consumes $\phi$ becomes more dominant.

>>>>

Traditionally, the transport of a given species with concentration $\phi$ convected by plug flow through a porous media without diffusion in one dimension (z) can be written as:

\begin{equation}
\epsilon v \frac{\partial \phi(z)}{\partial z}=S(z)
\end{equation}

@@

\begin{align}
\nonumber q(z, \xi) \cdot \frac{\partial \phi(z, \xi)}{\partial z}  &= S(z, \xi)   +
 \int [\phi(z, \hat \xi) - \phi(z, \xi)] h(z, \hat \xi \rightarrow \xi) d\hat \xi\\
\phi(z=z_{\Gamma},\xi)&=\phi_{\Gamma}(\xi).
\label{eq:ss2}
\end{align}

With $\phi_{\Gamma}(\xi)$ a boundary condition.
The left hand side and first term of the right hand side of the equation \ref{eq:ss2} are the convection per scale and the source per scale and present no major interaction between scales. However, the integral on the right hand side couples different scales and makes impossible to solve the problem for each scale independently. It should be pointed out that advection can be happening at different speeds ($\tfrac{\partial q}{\partial \xi} \neq 0$) and the source term can also be written independently for each pore-size.

@@

Find $\mathbf{f}\in X(\Omega)$ such that:
\begin{eqnarray}
\mathcal{L} \mathbf{f} = \mathbf{g} \quad \mbox{in} \ \Omega \label{eq:Problem} \\
\mathcal{B} \mathbf{f} = \mathbf{f}_{\Gamma} \quad \mbox{on} \ \Gamma \label{eq:Boundary}
\end{eqnarray}

@@

\begin{eqnarray}
\mathbf{f}&=&\phi_i \nonumber \\
\mathbf{g}&=&S_i \label{eq:problem} \\
\mathcal{L}\bullet&=& q \cdot \frac{\partial \bullet}{\partial z} - \int \bullet h(z, \hat \xi \rightarrow \xi) d\hat \xi +\bullet\int h(z, \hat \xi \rightarrow \xi) d\hat \xi \nonumber\\
\mathcal{B}\bullet&=&  \bullet
 \nonumber
\end{eqnarray}

@@

\begin{align}
\epsilon(\xi)=&A (\xi-\xi_0) (-\xi+\xi_f) \\
q({\xi}) =&V \epsilon(\xi) \xi^2\\
h(z, \hat \xi \rightarrow \xi) =& q(\hat \xi) q(\xi)/L\\
\phi(z, \xi) =& 10 sin(z) (1-M Log({\xi}/R))\\
\nonumber S(z, \xi)=& 10 A V cos(z) {\xi}^2 ({\xi}-\xi_0) (\xi_f-{\xi}) \left(1-M \text{Log}\left[\nicefrac{{\xi}}{R}\right]\right) +\\
\nonumber  &A^2 M V^2  (360 L)^{-1} {\xi}^2 ({\xi}-\xi_0) ({\xi}-\xi_f) sin(z) \times ... \\
\nonumber  &\big(-81 \xi_0^5+175 \xi_0^4 \xi_f-175 \xi_0 \xi_f^4+  ...\\
\nonumber &81 \xi_f^5+60 \xi_0^4 (3 \xi_0-5 \xi_f) \text{Log}\left[\frac{\xi_0}{{\xi}}\right]+ ...\\
& 60 (5 \xi_0-3 \xi_f) \xi_f^4 \text{Log}\left[\frac{\xi_f}{{\xi}}\right]\big)
\end{align}

@@

The solution is shown in Figure \ref{fig:solution}. As the liquid flows down the column $\phi$ increases due to reactions happening inside the pores. Despite the source term is very different for different pore sizes (negative for large pores and positive for small pores, see Figure \ref{fig:source}) the tracer $\phi$ is exchanged between different pores scales tending to level out the differences. However the increase is larger for the small pores at all the length of the column (z values). The source term (Figure \ref{fig:source}) behaves differently for different pore scales. Large amounts of $\phi$ are generated for pores smaller than 5 mm, while a sink is observed in the largest pores. This behaviour for the source term mimic two competing reactions. It models the case of a species that is a product of an heterogeneous reaction but is a reactant of a reaction taking place in the liquid bulk. Heterogeneous reactions depend strongly of the surface to volume ratio which is much larger for small pore scales. At the same time the reactant is consumed in the larger pores in this case as the bulk reaction that consumes $\phi$ becomes more dominant.
